---
runtime: shiny
title: "Monthly deaths in Slovenia since 1977"
output: html_document
---

```{r setup, include = TRUE, message = FALSE, warning = FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(readxl)
library(tidyr)
library(ggplot2)
library(shiny)
```

```{r echo = FALSE}
deathdata1 <- read.table("./data/umrljivost_1977_2018.txt", header = TRUE, sep = "\t", check.names = FALSE)
colnames(deathdata1)[1] <- "year"
deathdata1 <- pivot_longer(deathdata1, -year, names_to = "month", values_to = "deceased")
deathdata1$month <- sprintf("%02d", as.numeric(deathdata1$month))
deathdata1$year_month <- with(deathdata1, paste(year, month, "01", sep = "-"))
deathdata1$year_month <- as.Date(deathdata1$year_month, format = "%Y-%m-%d")

dd1 <- deathdata1[, c("year", "month", "deceased")]
# Because we have daily data for 2010-2020, we omit these from this set.
dd1 <- dd1[dd1$year < 2010, ]
```

```{r echo = FALSE}
dd2 <- read_excel("./data/umrljivost_update_2020_11_30.xls", sheet = "adapted")  # 2010-2020
dd2 <- as.data.frame(dd2)
dd2$value <- as.numeric(dd2$value)
dd2$datum <- as.Date(dd2$datum)

# TODO: because data in next chunk is for 2020 only, we need to omit it here
dd2 <- dd2[dd2$datum < as.Date("2020-01-01"), ]

dd2 <- data.frame(year = strftime(x = dd2$datum, format = "%Y"),
                  month = strftime(x = dd2$datum, format = "%m"),
                  deceased = dd2$value)
dd2 <- aggregate(deceased ~ year + month, data = dd2, FUN = sum)
```

```{r echo = FALSE}
dd3 <- read_excel("./data/umrljivost_updated_2020_12_09.xls", sheet = "adapted")
dd3 <- as.data.frame(dd3)
dd3$datum <- as.Date(dd3$datum)

# Data for December are not yet complete.
dd3 <- dd3[dd3$datum < as.Date("2020-12-01"), ]

dd3 <- data.frame(year = strftime(x = dd3$datum, format = "%Y"),
                  month = strftime(x = dd3$datum, format = "%m"),
                  deceased = dd3$value)
dd3 <- aggregate(deceased ~ year + month, data = dd3, FUN = sum)
```

```{r echo = FALSE}
dd <- rbind(dd1, dd2, dd3)
dd <- as.data.frame(dd)
dd$year <- as.numeric(dd$year)
```

```{r echo = FALSE, warning = FALSE, message = FALSE}
# Import daily data downloaded from SURS
# 
dbd <- read.table(file = "./data/surs_death_by_day_2000-2020-09.txt", 
                  header = TRUE, sep = "\t", nrows = 7719)  # dbd = death by day

colnames(dbd) <- c("year_month", "day", "count")
find.year.month <- "^(\\d{4})M(\\d{2}).*$"
dbd$year <- gsub(find.year.month, replacement = "\\1", x = dbd$year_month)
dbd$month <- gsub(find.year.month, replacement = "\\2", x = dbd$year_month)
dbd$datum <- as.Date(paste(dbd$year, dbd$month, dbd$day, sep = "-"))

dbd$count <- as.numeric(as.character(dbd$count))

# Use a fake year to align the data.
dbd$fake_datum <- as.Date(paste(1, dbd$month, dbd$day, sep = "-"))

dbd$year_month <- NULL

dbd <- dbd[!is.na(dbd$datum), ]

dbd.download <- dbd[, c("year", "month", "day", "count")]
```

```{r echo = FALSE}
sliderInput(
  inputId = "sel_year",
  min = min(dd$year),
  max = max(dd$year),
  step = 1,
  label = "Highlight year",
  value = 2020,
  width = "100%",
  sep = ""
)

sliderInput(
  inputId = "sel_background",
  min = min(dd$year),
  max = max(dd$year),
  step = 1,
  label = "Select background range",
  value = c(min(dd$year), max(dd$year)),
  width = "100%",
  sep = ""
)
```

```{r echo = FALSE}
renderPlot({
  subdd <- dd[dd$year >= input$sel_background[1] & 
                dd$year <= input$sel_background[2], ]
  
  ggplot(subdd, aes(x = month, y = deceased, group = year)) +
    theme_bw() +
    xlab("Month") + ylab("Number of deaths per month") +
    geom_violin(aes(group = month), alpha = 0.5) +
    geom_line(alpha = 0.5) +
    geom_line(data = dd[dd$year == input$sel_year, ], color = "red", size = 1)
}, res = 100)
```

```{r echo = FALSE}
downloadLink(
  outputId = "download_data",
  label = "Download monthly data"
)

output$download_data <- downloadHandler(
  filename = function() {
    sprintf("monthly_deaths_slovenia_%s_%s.csv", min(dd$year), max(dd$year))
  },
  content = function(con) {
    write.table(dd, file = con, quote = FALSE, row.names = FALSE, sep = ",")
  }
)
```

```{r echo = FALSE, fig.width = 5, fig.height = 3}
renderPlot({
  ggplot(dbd, aes(x = fake_datum, y = count, group = year)) +
    theme_bw() +
    xlab("Datum") +
    ylab("Number of deaths per day") +
    scale_x_date(date_labels = "%b%d", date_breaks = "1 month") +
    geom_line(alpha = 0.4) +
    geom_point(alpha = 0.4) +
    geom_point(data = function(x) x[x$year == input$sel_year, ], color = "red") +
    geom_line(data = function(x) x[x$year == input$sel_year, ], color = "red")
}, res = 100)
```

```{r echo = FALSE}
downloadLink(
  outputId = "download_daily_data",
  label = "Download daily data"
)

output$download_daily_data <- downloadHandler(
  filename = function() {
    "daily_death_data_slovenia_since_2000.csv"
  },
  content = function(con) {
    write.table(dbd.download, file = con, quote = FALSE, row.names = FALSE, sep = ",")
  }
)
```

### Compare years by month(s)
```{r echo = FALSE}
selectInput(
  inputId = "sel_month",
  label = "Select month",
  choices = sort(unique(dd$month)),
  selected = "01",
  multiple = TRUE
)
```

```{r echo = FALSE, warning = FALSE}
renderPlot({
  if (is.null(input$sel_month)) return(NULL)
  
  ggplot(dd[dd$month %in% input$sel_month, ], aes(x = year, y = deceased, fill = month)) +
    theme_bw() +
    theme(legend.position = "top") +
    xlab("Years") +
    ylab("Number of deaths per month") +
    scale_fill_brewer(palette = "Paired") +
    geom_bar(stat = "identity", position = "dodge")
}, res = 100)
```

#### Daily data for all years
```{r echo = FALSE, warning = FALSE, message = FALSE}
renderPlot({
  ggplot(dbd, aes(x = fake_datum, y = count, group = year)) +
    theme_bw() +
    xlab("Datum") + ylab("Deaths per day") +
    geom_line(size = 0.5) +
    geom_point() +
    geom_smooth(method = "loess", color = "red", se = FALSE, span = 0.1) +
    facet_wrap(~ year, ncol = 3)
}, res = 100, width = 1000, height = 1500)
```
(red line is a loess smoothed line with span of 0.1)

[Matjaž Jeran provided data](https://sledilnik.slack.com/archives/CV93RSH17/p1587931284356400) on number of deaths by year-month for Slovenia for era 1977-2009. [Andrej Viršček provided daily data](https://sledilnik.slack.com/archives/CV93RSH17/p1588680372152800?thread_ts=1587931284.356400&cid=CV93RSH17) for 2010-2020 (source Ministry of the interior). Data comes from Ministry of interior and includes also persons who are not citizens but have died in Slovenia. Originally the data is day-based but is summarized to months to make it compatible with historic data.

Daily data downloaded from [stat.si](https://pxweb.stat.si/SiStatData/pxweb/sl/Data/-/05L1018S.px) as a "relational" file. Note that _recent daily data may be provisional and subject to change_. It includes persons with permanent residence in Slovenia, possibly also persons who have died when abroad. Last accessed on 2020-11-28.

App by Roman Luštrik, last updated on `r Sys.Date()`. Contact me on twitter @romunov.